---
title: "RNA-seq - LRT example"
author: "Catherine Chahrour"
date: "`r format(Sys.time(), '%d %B %Y')`"
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-location: left
    embed-resources: true
    theme: cosmo 
    highlight-style: github-dark
    number-sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=F}
# Set global chunk options and load required packages quietly
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.height = 6,
  fig.width = 8,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.lazy = FALSE,
  results = 'asis' 
)

# Suppress messages while loading libraries for cleaner output
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(circlize))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(qvalue))
suppressPackageStartupMessages(library(limma))

# Set working directory (adjust as needed)
cwd <- getwd()
setwd(dir = cwd)
```

RNA-seq Data from:

Meaker GA, Nicholls M, Chahrour C, Hsu I, Smith A, Bozhilov Y, Leung M, Vassort H, Olender L, Beaven O, Huang X, Brown EJ, Vanden Bempt M, Khoo HM, Bhadury J, Milne TA, Wilkinson AC. A genome-wide screen identifies Runx2 as a novel regulator of hematopoietic stem cell expansion and T-cell commitment. Blood. 2025 Dec 25;146(26):3188-3200. 
doi: 10.1182/blood.2025029115. PubMed PMID: 40961240.

This report uses the HPC raw count matrix from GSE288012 to demonstrate a DESeq2 likelihood ratio test (LRT). The LRT asks whether genotype explains expression variation overall across the `hom`, `het`, and `wt` groups, rather than testing only one pairwise contrast.


```{r}
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE288012&format=file&file=GSE288012%5Fraw%5Fcounts%5FHPC%2Etsv%2Egz"
dest <- "GSE288012_raw_counts_HPC.tsv.gz"

download.file(url, destfile = dest, mode = "wb")
df <- read.delim(gzfile(dest), header = TRUE, skip = 1, check.names = FALSE)
```

# Sample Information

Sample metadata are reconstructed from the count matrix column names. The sample labels are split into the cell population, genotype, and replicate fields, and all three genotype groups are retained for the LRT.

```{r}
# Load sample metadata and convert character and numeric columns to factors
sample_info <- data.frame(sample = colnames(df)) %>%
  filter(startsWith(sample, "seqnado_output")) %>% 
  mutate(sample = gsub("(.+aligned/)(.+)(.bam)", "\\2", sample)) %>% 
  separate(col = sample, into = c("cell", "genotype", "replicate"), remove = F) %>% 
  mutate(across(where(is.character), as.factor)) %>%
  mutate(across(where(is.numeric), as.factor)) 

sample_info$genotype <- relevel(sample_info$genotype, ref = "wt")
# Show sample metadata table
kable(sample_info, row.names = F, "simple")
```

```{r}
# Define consistent color palettes for time and treatment variables
gene_col <- brewer.pal(9, "Reds")[c(5,7,9)] %>% 
  setNames(levels(sample_info$genotype))


# Define heatmap row annotation using sample metadata
row_ha <- rowAnnotation(
  genotype = sample_info$genotype,
  col = list(
    genotype = gene_col
  ),
  show_annotation_name = FALSE
)

# Define heatmap column annotation using sample metadata
col_ha <- HeatmapAnnotation(
  genotype = sample_info$genotype,
  col = list(
    genotype = gene_col
  ),
  show_annotation_name = FALSE
)
```

# Counts from SeqNado

The raw count table is converted into the matrix format expected by DESeq2: genes in rows and samples in columns. Columns are renamed to match the cleaned sample metadata, all selected genotype samples are retained, and genes with zero counts across all samples are removed.

```{r}
# Load and clean count matrix from featureCounts output
counts <- df %>%
  rename_with(~ gsub("(.+aligned/)(.+)(.bam)", "\\2", .), everything()) %>% 
  column_to_rownames(var = "Geneid") %>%
  dplyr::select(all_of(sample_info$sample)) %>%
  filter(rowSums(.) > 0) %>% 
  as.matrix()
```

# DE analysis

The full model includes genotype (`~ genotype`), while the reduced model contains only the intercept (`~ 1`). The LRT compares these models gene by gene, identifying genes for which adding genotype significantly improves model fit.

```{r}
# Create DESeq2 dataset using full model for genotype
dds <- DESeqDataSetFromMatrix(
  countData = counts, 
  colData = sample_info,
  design = ~ genotype
)

# Fit model using likelihood ratio test (LRT), removing intersect in reduced model
dds <- DESeq(dds, test="LRT", reduced = ~ 1)
```

Normalisation was performed in DESeq2 and data vst transformed to allow for visualisation.

## RLE

Relative log expression (RLE) plots provide a sample-level quality control check after variance-stabilizing transformation. Well-normalized samples should have broadly similar RLE distributions centered around zero; large shifts or unusually wide boxes can indicate sample-specific bias or outliers.

```{r}
# Variance stabilizing transformation followed by batch correction
vst_counts <- dds %>% 
  vst() %>% 
  assay()

# Compute RLE by subtracting row medians
rle_matrix <- vst_counts - rowMedians(vst_counts)

# Reshape and merge with sample metadata
rle_df <- rle_matrix %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  melt(id.vars = "gene", variable.name = "sample", value.name = "RLE")%>% 
  inner_join(sample_info, by = join_by(sample))

# Plot RLE boxplot per sample, colored by genotype
ggplot(rle_df, aes(x = sample, y = RLE, fill = genotype)) +
  geom_boxplot(outlier.size = 0.5) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = gene_col) +
  labs(
    title = "Relative Log Expression (RLE) Plot", 
    y = "RLE (log2 scale)", 
    x = "Sample"
  )
```

## Sample distance matrix

The sample distance matrix compares samples using the variance-stabilized expression values. Samples with similar global expression profiles cluster together, which helps check whether biological replicates and genotype groups behave as expected.

```{r}
# Compute sample-to-sample Euclidean distance matrix
sampleDists <- vst_counts %>% 
  t() %>% 
  dist() %>% 
  as.matrix()

# Plot sample distance matrix as heatmap
set.seed(123)
HM_sampleDists <- Heatmap(
  sampleDists, 
  top_annotation = col_ha,
  left_annotation = row_ha,
  row_title = NULL,
  show_row_names = TRUE,
  cluster_rows = TRUE,
  column_title = NULL, 
  show_column_names = FALSE,
  cluster_columns = TRUE,
  show_heatmap_legend = FALSE,
  col = colorRamp2(c(0, max(sampleDists)), c("black", "white"))
)

draw(HM_sampleDists, merge_legend = TRUE)
```

## Overview by PCA {.tabset}

Principal component analysis (PCA) summarizes the main axes of variation across samples. In this plot, each point is a sample, colored by genotype and labeled by replicate, so genotype separation and replicate consistency can be assessed visually.

```{r}
# Perform PCA and calculate percent variance per PC
pca_res <- vst_counts %>% 
  t() %>% 
  prcomp()

percentVar <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), digits = 2)

# Merge with metadata
pca_data <- pca_res$x %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>%
  inner_join(sample_info, by = join_by(sample))

# PCA plot with sample labels
ggplot(pca_data, aes(PC1, PC2, color = genotype)) +
  geom_point(size = 4) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  geom_label_repel(
    aes(label = replicate), 
    color = "black", 
    fill = "white",
    size = 2,
    max.overlaps = 20,
    show.legend = FALSE) +
  scale_color_manual(values = gene_col) +
  ggtitle("PCA - PC1 vs PC2") +
  theme_light() 
```

## PCA correlate with covariates

This heatmap checks which retained sample metadata variables are most strongly associated with the leading principal components. Strong correlations can help identify whether genotype or replicate structure is driving the dominant variation in the dataset.

```{r}
# Convert metadata to numeric factors for correlation with PCs
metadata_for_cor <- colData(dds) %>%
  as.data.frame() %>% 
  dplyr::select(-c(sample, cell)) %>% 
  mutate_if(is.character, as.factor) %>% 
  mutate_if(is.factor, as.integer)

# Correlation between metadata and first 8 PCs using Spearman method
pca_meta_corr <- abs(cor(
  metadata_for_cor, 
  pca_data[, paste0("PC", 1:8)], 
  method = "spearman"))

# Plot correlation heatmap
heatmap <- Heatmap(
  pca_meta_corr, 
  name = "Correlation", 
  col = colorRamp2(
    seq(0, 1, length = 9), 
    brewer.pal(n = 9, name = "Blues")), 
  cluster_rows = FALSE,
  cluster_columns = FALSE, 
  width = ncol(pca_meta_corr)*unit(1, "cm"), 
  height = nrow(pca_meta_corr)*unit(1, "cm"),
  heatmap_legend_param = list(
    title = "Correlation",
    title_position = "topcenter",
    direction = "horizontal")
)

draw(heatmap, heatmap_legend_side = "bottom", annotation_legend_side = "right")
```

# Heatmap

padj < 0.05

The LRT result table is used to identify genes with a significant overall genotype effect. The heatmap then shows the expression pattern of genes significant at `padj < 0.05`, using row-scaled normalized counts so that relative expression differences across the `hom`, `het`, and `wt` groups are easier to compare.

```{r}
normCounts <- counts(dds, normalized = T) %>%
  as.data.frame() %>%
  rownames_to_column("gene")

mean_normCounts <- normCounts %>%
  as.data.frame() %>%
  gather(sample, value, -gene) %>%
  left_join(sample_info, by = c(sample = "sample")) %>%
  group_by(gene, cell) %>%
  summarise(mean_value = mean(value), .groups = "rowwise") %>%
  spread(key = cell, value = mean_value, sep = "_mean_")

results <- results(dds) %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  inner_join(mean_normCounts, by = "gene") %>%
  filter(!is.na(padj)) %>%
  filter(!is.na(pvalue)) %>%
  arrange(padj)
```


```{r}
results_sig <- results %>%
  filter(padj < 0.05) %>%
  arrange(padj) %>%
  head(5000)

matrix <- normCounts %>%
  as.data.frame() %>%
  filter(gene %in% results_sig$gene) %>%
  column_to_rownames("gene") %>%
  as.matrix() %>%
  t() %>%
  scale() %>%
  t()

groups <- as.character(sample_info$genotype)
col_fun <- colorRamp2(seq(1.5, -1.5, length = 11), brewer.pal(n = 11, name = "RdBu"))

set.seed(123)
DEheatmap <- Heatmap(
  matrix,
  top_annotation = col_ha,
  cluster_rows = T,
  show_row_names = F,
  row_title = "genes",
  show_row_dend = T,
  show_column_names = FALSE,
  cluster_columns = F,
  column_split = groups,
  # row_km = 3,
  col = col_fun,
  use_raster = T,
  row_title_gp = gpar(fontsize = 16),
  column_title_gp = gpar(fontsize = 16, fontface = "bold"),
  heatmap_legend_param = list(
    title = "Row z-score",
    title_gp = gpar(fontsize = 16),
    labels_gp = gpar(fontsize = 16),
    title_position = "topcenter",
    direction = "horizontal"))

ht = draw(DEheatmap, heatmap_legend_side = "bottom", annotation_legend_side = "right")

# pdf(file=paste0(fig_dir, "/heatmap.pdf"), width=8, height=6, pointsize = 16)
# draw(ht, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
# dev.off()

```

# Number of significant genes

Because the LRT tests the genotype term as a whole, this summary counts significant genes using adjusted p-value thresholds only. It does not apply log2 fold-change cutoffs or assign genes to up/down categories, since a multi-level LRT does not have a single direction of effect.

```{r}
results <- results(dds) %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  filter(!is.na(padj)) %>%
  filter(!is.na(pvalue)) %>%
  arrange(padj)

padj_thresholds <- c(0.05, 0.01, 0.001)

DEsummary <- data.frame(
  padj_threshold = factor(
    paste0("padj < ", padj_thresholds),
    levels = paste0("padj < ", padj_thresholds)
  ),
  count = sapply(
    padj_thresholds,
    function(padj_cutoff) sum(results$padj < padj_cutoff, na.rm = TRUE)
  )
)

ggplot(DEsummary, aes(x = padj_threshold, y = count)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = count), vjust = -0.3, size = 5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  ggtitle("Significant genes from LRT") +
  labs(
    x = "Adjusted p-value threshold",
    y = "Number of significant genes"
  ) +
  theme_light() +
  theme(text = element_text(size = 15))
```


# Pairwise comparisons {.tabset}

The overall LRT identifies genes with evidence of genotype-associated expression changes. These follow-up coefficient plots provide exploratory pairwise views of the fitted genotype effects, helping show which genotype comparisons may contribute to the overall signal.

```{r}
colors <- brewer.pal(5, "RdBu")[c(5,1)] %>% 
  append("grey") %>% 
  set_names(c("Down", "Up", "NotSig"))

res_names <- resultsNames(dds)[-1]

DE_results_pairwise <- list()

for(comp in res_names) {
  cat("## ", comp, "\n \n")
  
  res <- results(dds, name = comp, test="Wald") %>% 
   as.data.frame() %>%
    rownames_to_column("gene") %>% 
    arrange(padj) %>% 
    filter(!is.na(padj))
  
  pval <- ggplot(res, aes(pvalue)) +
    geom_histogram(fill = "lightblue", binwidth = 0.01) +
    theme_light() +
    theme(text = element_text(size = 15)) +
    ggtitle(paste("P-value distribution: ", comp)) 
  print(pval)
  
  res_filtered <- res %>% 
    filter(padj < 0.05)
  
  res_list <- list(res_filtered)
  names(res_list) <- comp
  DE_results_pairwise <- append(DE_results_pairwise, res_list)

  res4plot <- res %>%
    dplyr::select(gene, log2FoldChange, padj) %>%
    arrange(padj) %>%
    mutate(
      DE = case_when(
        padj < 0.05 & log2FoldChange < 0 ~ "Down",
        padj < 0.05 & log2FoldChange >= 0 ~ "Up",
        TRUE ~ "NotSig"
      ),
      label = if_else(gene %in% union(
        head(filter(., DE == "Up")$gene, 5),
        head(filter(., DE == "Down")$gene, 5)
      ), gene, as.character(NA))
    )

  maxLFC <- max(abs(res4plot$log2FoldChange))
  maxpadj <- max(-log10(res4plot$padj), na.rm = TRUE)
  
  
  g <- ggplot(data=res4plot, aes(x=log2FoldChange, y=-log10(padj), col=DE, label=label)) +
    geom_point(size=1) +
    theme_minimal() +
    geom_label_repel(aes(label = label), show.legend = F, max.overlaps = 20) +
    scale_colour_manual(values = colors) +
    geom_vline(xintercept=0, linetype='dotted', col="black") +
    geom_hline(yintercept=-log10(0.05), linetype='dotted', col="black") +
    xlab("log2 fold change") +
    ylab("-log10 adjusted p-value") +
    xlim(-maxLFC, maxLFC) +
    ylim(0, maxpadj) +
    ggtitle(comp) +
    theme(text = element_text(size = 15))
  # outname <- paste0("plots/volcano_", comp,".pdf")
  # ggsave( outname, g, width = 8, height = 6)
  print(g)
  
  cat(" \n \n")
}
```

# Overlapping genes in comparisons

This Venn diagram compares the significant gene sets from the follow-up genotype coefficients. It helps show whether the `hom` and `het` comparisons highlight largely shared or distinct subsets of genes relative to `wt`.

```{r}
library(venn)
venn::venn(
  x = list(
    "RUNX2 hom" = DE_results_pairwise$genotype_hom_vs_wt$gene,
    "RUNX2 het" = DE_results_pairwise$genotype_het_vs_wt$gene
  ),
  zcolor = gene_col[c("het", "hom")],
  opacity = 0.5,
  ilcs = 1,
  ilabels = "counts",
  box = F,
)
```


# Session Info

The session information records the R version and package versions used to render the report. This is useful for reproducibility, especially for Bioconductor workflows where package versions can affect results.

```{r results='markup'}
sessionInfo()
```
