---
title: "RNA-seq - LRT example"
author: "Catherine Chahrour"
date: "`r format(Sys.time(), '%d %B %Y')`"
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-location: left
    embed-resources: true
    theme: cosmo 
    highlight-style: github-dark
    number-sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=F}
# Set global chunk options and load required packages quietly
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.height = 6,
  fig.width = 8,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.lazy = FALSE,
  results = 'asis' 
)

# Suppress messages while loading libraries for cleaner output
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(circlize))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(qvalue))
suppressPackageStartupMessages(library(limma))

# Set working directory (adjust as needed)
# setwd(dir = "/path/to/working_directory")
```

RNA-seq Data from:

Meaker GA, Nicholls M, Chahrour C, Hsu I, Smith A, Bozhilov Y, Leung M, Vassort H, Olender L, Beaven O, Huang X, Brown EJ, Vanden Bempt M, Khoo HM, Bhadury J, Milne TA, Wilkinson AC. A genome-wide screen identifies Runx2 as a novel regulator of hematopoietic stem cell expansion and T-cell commitment. Blood. 2025 Dec 25;146(26):3188-3200. 
doi: 10.1182/blood.2025029115. PubMed PMID: 40961240.

```{r}
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE288012&format=file&file=GSE288012%5Fraw%5Fcounts%5FHPC%2Etsv%2Egz"
dest <- "GSE288012_raw_counts_HPC.tsv.gz"

download.file(url, destfile = dest, mode = "wb")
df <- read.delim(gzfile(dest), header = TRUE, skip = 1, check.names = FALSE)
```


# Sample Information

```{r}
# Load sample metadata and convert character and numeric columns to factors
sample_info <- data.frame(sample = colnames(df)) %>%
  filter(startsWith(sample, "seqnado_output")) %>% 
  mutate(sample = gsub("(.+aligned/)(.+)(.bam)", "\\2", sample)) %>% 
  separate(col = sample, into = c("cell", "genotype", "replicate"), remove = F) %>% 
  filter(genotype != "hom") %>% # Here we only wnt the wt and het samples
  mutate(across(where(is.character), as.factor)) %>%
  mutate(across(where(is.numeric), as.factor)) 

sample_info$genotype <- relevel(sample_info$genotype, ref = "wt")
# Show sample metadata table
kable(sample_info, row.names = F, "simple")
```

```{r}
# Define consistent color palettes for time and treatment variables
gene_col <- brewer.pal(9, "Reds")[c(5,9)] %>% 
  setNames(levels(sample_info$genotype))


# Define heatmap row annotation using sample metadata
row_ha <- rowAnnotation(
  genotype = sample_info$genotype,
  col = list(
    genotype = gene_col
  ),
  show_annotation_name = FALSE
)

# Define heatmap column annotation using sample metadata
col_ha <- HeatmapAnnotation(
  genotype = sample_info$genotype,
  col = list(
    genotype = gene_col
  ),
  show_annotation_name = FALSE
)
```

# Counts from SeqNado

```{r}
# Load and clean count matrix from featureCounts output
counts <- df %>%
  rename_with(~ gsub("(.+aligned/)(.+)(.bam)", "\\2", .), everything()) %>% 
  column_to_rownames(var = "Geneid") %>%
  dplyr::select(all_of(sample_info$sample)) %>%
  filter(rowSums(.) > 0) %>% 
  as.matrix()
```

# DE analysis

```{r}
# Create DESeq2 dataset using full model for genotype
dds <- DESeqDataSetFromMatrix(
  countData = counts, 
  colData = sample_info,
  design = ~ genotype
)

# Fit model using Wald test
dds <- DESeq(dds, test = "Wald")
```

Normalisation was performed in DESeq2 and data vst transformed to allow for visualisation.

## RLE

```{r}
# Variance stabilizing transformation followed by batch correction
vst_counts <- dds %>% 
  vst() %>% 
  assay()

# Compute RLE by subtracting row medians
rle_matrix <- vst_counts - rowMedians(vst_counts)

# Reshape and merge with sample metadata
rle_df <- rle_matrix %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  melt(id.vars = "gene", variable.name = "sample", value.name = "RLE")%>% 
  inner_join(sample_info, by = join_by(sample))

# Plot RLE boxplot per sample, colored by genotype
ggplot(rle_df, aes(x = sample, y = RLE, fill = genotype)) +
  geom_boxplot(outlier.size = 0.5) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = gene_col) +
  labs(
    title = "Relative Log Expression (RLE) Plot", 
    y = "RLE (log2 scale)", 
    x = "Sample"
  )
```

## Sample distance matrix

```{r}
# Compute sample-to-sample Euclidean distance matrix
sampleDists <- vst_counts %>% 
  t() %>% 
  dist() %>% 
  as.matrix()

# Plot sample distance matrix as heatmap
set.seed(123)
HM_sampleDists <- Heatmap(
  sampleDists, 
  top_annotation = col_ha,
  left_annotation = row_ha,
  row_title = NULL,
  show_row_names = TRUE,
  cluster_rows = TRUE,
  column_title = NULL, 
  show_column_names = FALSE,
  cluster_columns = TRUE,
  show_heatmap_legend = FALSE,
  col = colorRamp2(c(0, max(sampleDists)), c("black", "white"))
)

draw(HM_sampleDists, merge_legend = TRUE)
```

## Overview by PCA {.tabset}

```{r}
# Perform PCA and calculate percent variance per PC
pca_res <- vst_counts %>% 
  t() %>% 
  prcomp()

percentVar <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), digits = 2)

# Merge with metadata
pca_data <- pca_res$x %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>%
  inner_join(sample_info, by = join_by(sample))

# PCA plot with sample labels
ggplot(pca_data, aes(PC1, PC2, color = genotype)) +
  geom_point(size = 4) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  geom_label_repel(
    aes(label = replicate), 
    color = "black", 
    fill = "white",
    size = 2,
    max.overlaps = 20,
    show.legend = FALSE) +
  scale_color_manual(values = gene_col) +
  ggtitle("PCA - PC1 vs PC2") +
  theme_light() 
```

## PCA correlate with covariates

```{r}
# Convert metadata to numeric factors for correlation with PCs
metadata_for_cor <- colData(dds) %>%
  as.data.frame() %>% 
  dplyr::select(-c(sample, cell)) %>% 
  mutate_if(is.character, as.factor) %>% 
  mutate_if(is.factor, as.integer)

# Correlation between metadata and first 8 PCs using Spearman method
pca_meta_corr <- abs(cor(
  metadata_for_cor, 
  pca_data[, paste0("PC", 1:6)], 
  method = "spearman"))

# Plot correlation heatmap
heatmap <- Heatmap(
  pca_meta_corr, 
  name = "Correlation", 
  col = colorRamp2(
    seq(0, 1, length = 9), 
    brewer.pal(n = 9, name = "Blues")), 
  cluster_rows = FALSE,
  cluster_columns = FALSE, 
  width = ncol(pca_meta_corr)*unit(1, "cm"), 
  height = nrow(pca_meta_corr)*unit(1, "cm"),
  heatmap_legend_param = list(
    title = "Correlation",
    title_position = "topcenter",
    direction = "horizontal")
)

draw(heatmap, heatmap_legend_side = "bottom", annotation_legend_side = "right")
```


# DE results

```{r}
normCounts <- counts(dds, normalized = T) %>%
  as.data.frame() %>%
  rownames_to_column("gene")

mean_normCounts <- normCounts %>%
  as.data.frame() %>%
  gather(sample, value, -gene) %>% 
  left_join(sample_info, by = c(sample = "sample")) %>% 
  group_by(gene, cell) %>% 
  summarise(mean_value = mean(value), .groups = "rowwise") %>%
  spread(key = cell, value = mean_value, sep = "_mean_")

results <- results(dds) %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  inner_join(mean_normCounts, by = "gene") %>%
  filter(!is.na(padj)) %>%
  filter(!is.na(pvalue)) %>%
  arrange(padj)
```

## Histogram of p-values {.tabset}

```{r}

p_val <- ggplot(results, aes(pvalue)) +
  geom_histogram(binwidth = 0.01, fill="lightblue") +
  scale_x_continuous(expand = c(0, 0), ) +
  scale_y_continuous(expand = c(0, 0), ) +
  ggtitle(paste0("Distribution of p-values")) +
  theme_minimal()

print(p_val)


# ggsave(
#   paste0(fig_dir, "/p_val_distribution.pdf"),
#   plot = p_val, device = cairo_pdf, dpi = 1200, width = 6, height = 6)
```

## Number of Diff_genes genes

```{r}
summary_fn <- function(df, pvals, fc_thresholds) {
  summary_df <- data.frame()
  
  for (pval in pvals) {
    for (i in seq_along(fc_thresholds)) {
      if (i == length(fc_thresholds)) {
        count_pos <- sum(df$padj < pval & df$log2FoldChange >= fc_thresholds[i], na.rm = TRUE)
        count_neg <- sum(df$padj < pval & df$log2FoldChange <= -fc_thresholds[i], na.rm = TRUE)
      } else {
        count_pos <- sum(df$padj < pval & df$log2FoldChange >= fc_thresholds[i] & df$log2FoldChange < fc_thresholds[i + 1], na.rm = TRUE)
        count_neg <- sum(df$padj < pval & df$log2FoldChange <= -fc_thresholds[i] & df$log2FoldChange > -fc_thresholds[i + 1], na.rm = TRUE)
      }
      summary_df[1, paste0('fc_', fc_thresholds[i], '_pos_', pval)] <- count_pos
      summary_df[1, paste0('fc_', fc_thresholds[i], '_neg_', pval)] <- -count_neg  
    }
  }
  
  summary_df %>%
    pivot_longer(cols = starts_with("fc_"), values_to = "count") %>%
    separate(col = name, into = c("fc", "fc_value", "direction", "pvalue"), sep = "_") %>%
    mutate(fc_value = paste0("fc_", fc_value),
           fc_value = factor(fc_value, levels = rev(unique(fc_value))),
           pvalue = factor(pvalue, levels = unique(pvalue)))
}

# Define p-values and FC thresholds
pvalues <- c(0.05, 0.01, 0.001)
fc_thresholds <- c(0:3)

# run our little function
diff_summary <- summary_fn(results, pvalues, fc_thresholds)

# Just making the colours more contrasty
fc_col <- brewer.pal(n = length(fc_thresholds) + 2, "Purples")[-c(1:2)] %>% 
  set_names(paste0("fc_", fc_thresholds))
res_names <- resultsNames(dds)[-1]
# Plot the summary
ggplot(diff_summary, aes(x = pvalue, y = count, fill = fc_value)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_hline(yintercept = 0, color = "darkgrey") +
  ggtitle(paste("Diff_genes -", res_names)) +
  scale_fill_manual(values = fc_col) +
  theme_bw() +
  labs(x = "P-value Threshold", y = "Count")
```

# Volcano

```{r}
colors <- brewer.pal(5, "RdBu")[c(5,1)] %>%
  append("grey") %>%
  set_names(c("Down", "Up", "NotSig"))


res4plot <- results %>%
  dplyr::select(log2FoldChange, padj, gene) %>%
  mutate(padj = padj +1e-300) %>% 
  arrange(padj) %>%
  mutate(
    DE = case_when(
      padj < 0.05 & log2FoldChange < 0 ~ "Down",
      padj < 0.05 & log2FoldChange > 0 ~ "Up",
      TRUE ~ "NotSig"
    ),
    label = if_else(gene %in% union(
      head(filter(., DE == "Up")$gene, 5),
      head(filter(., DE == "Down")$gene, 5)
    ), gene, as.character(NA))
  ) 
maxLFC <- max(abs(res4plot$log2FoldChange))
maxpadj <- max(-log10(res4plot$padj), na.rm = TRUE)
maxpadj <- max(maxpadj, -log10(0.05))

g <- ggplot(
  data=res4plot,
  aes(
    x=log2FoldChange,
    y=-log10(padj),
    col=DE,
    label=label)) +
  geom_point(size = 1, alpha = 0.5) +
  theme_minimal() +
  geom_label_repel(aes(label = label), show.legend = T) +
  scale_colour_manual(values = colors) +
  geom_vline(xintercept=0, linetype='dotted', col="black") +
  geom_hline(yintercept=-log10(0.05), linetype='dotted', col="black") +
  xlab("log2 fold change") +
  ylab("-log10 adjusted p-value") +
  theme(axis.title.x = element_text(colour="black", size=14),
        axis.title.y = element_text(colour="black", size=14))+
  xlim(-maxLFC, maxLFC) +
  ylim(0, maxpadj) +
  ggtitle(res_names)

# ggsave(
#   filename = paste0(de_dir, "/volcano_", res_names, ".pdf"),
#   plot = g,
#   dpi = 900,
#   create.dir = T)

print(g)
```

# Heatmap

padj < 0.05

```{r}
results_sig <- results %>%
  filter(padj < 0.05) %>%
  arrange(padj) %>%
  head(5000)

matrix <- normCounts %>%
  as.data.frame() %>%
  filter(gene %in% results_sig$gene) %>%
  column_to_rownames("gene") %>%
  as.matrix() %>%
  t() %>%
  scale() %>%
  t()

groups <- as.character(sample_info$genotype)
col_fun <- colorRamp2(seq(1.5, -1.5, length = 11), brewer.pal(n = 11, name = "RdBu"))

set.seed(123)
DEheatmap <- Heatmap(
  matrix,
  top_annotation = col_ha,
  cluster_rows = T,
  show_row_names = F,
  row_title = "genes",
  show_row_dend = T,
  show_column_names = FALSE,
  cluster_columns = T,
  column_split = groups,
  col = col_fun,
  use_raster = T,
  row_title_gp = gpar(fontsize = 16),
  column_title_gp = gpar(fontsize = 16, fontface = "bold"),
  heatmap_legend_param = list(
    title = "Row z-score",
    title_gp = gpar(fontsize = 16),
    labels_gp = gpar(fontsize = 16),
    title_position = "topcenter",
    direction = "horizontal"))

ht = draw(DEheatmap, heatmap_legend_side = "bottom", annotation_legend_side = "right")

# pdf(file=paste0(fig_dir, "/heatmap.pdf"), width=8, height=6, pointsize = 16)
# draw(ht, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
# dev.off()

```

# Session Info

```{r results='markup'}
sessionInfo()
```

